---
title: "Is happiness U-shaped?"
author: "Andi Fugard (almost@gmail.com, @[inductivestep](https://twitter.com/InductiveStep))"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    self_contained: no
    toc: yes
    toc_float: yes
    collapsed: false
---

# Introduction

This is **work in progress**, in response to David G. Blanchflower's paper [Is Happiness U-shaped Everywhere?](https://www.nber.org/papers/w26641).

I'm curious to know:

* The shape of happiness in the latest round of the [European Social Survey](https://www.europeansocialsurvey.org/).
* How the magnitude of any effect for age compares with other obvious predictors of wellbeing.
  
Load packages:

```{r}
library(tidyverse)
library(haven)
library(srvyr)
library(quantreg)
library(survey)
library(broom)
```

Read in the ESS data, [integrated file, edition 2.0 round 9](https://www.europeansocialsurvey.org/download.html?file=ESS9e02&c=&y=2018):

```{r}
ESS9e02 <- read_sav("ESS9e02.sav")
```

Have a look:

```{r eval = FALSE}
View(ESS9e02)
```



# Finding the relevant variables

The variable docs are [available here](http://nesstar.ess.nsd.uib.no/webview/).

* `happy`: "Taking all things together, how happy would you say you are?"
* `gndr`: Gender (only male or female)
* `agea`: Age in years
* `cntry`: Country of respondent
* `pspwght`: Post-stratification weights ([Docs](https://www.europeansocialsurvey.org/docs/methodology/ESS_weighting_data_1.pdf))

```{r}
dat <- ESS9e02 %>%
  select(happy,
         gndr,
         agea,
         cntry,
         pspwght)
```

Let's see how many observations there are by country.

```{r}
raw_obs <- dat %>%
  group_by(cntry) %>%
  summarise(n_raw = n())
```

Make a survey object, using the `srvyr` wrapper around the `survey` package:

```{r}
dat_svy <- dat %>%
   as_survey_design(weights = pspwght)
```



Let's first plot the mean happiness by age group (weighted appropriately), to get a sense of what might be going on.

First, how does age look?

```{r}
summary(dat$agea)
```

As a first try, let's group into fives:

```{r}
table(cut(dat$agea, seq(15,90,5), include.lowest = T))
```
Looks like it should be enough.




```{r}
mean_happy <- dat_svy %>%
  mutate(age_grp = cut(agea, seq(15,90,5))) %>%
  select(age_grp, cntry, happy) %>%
  na.omit() %>%
  group_by(cntry, age_grp) %>%
  summarise(Happiness = survey_mean(happy))
mean_happy
```


```{r}
age_group_names <- levels(mean_happy$age_grp)
age_group_names
```

```{r}
whichLabels <- c(2,
                 floor(length(age_group_names)/2 + 1),
                 length(age_group_names) - 1)

```


```{r fig.height=15, fig.width=7}
mean_happy %>%
  ggplot(aes(x = age_grp, y = Happiness,
             ymin = Happiness - 1.96*Happiness_se,
             ymax = Happiness + 1.96*Happiness_se)) +
  geom_pointrange() +
  geom_line(group = 1) +
  scale_x_discrete(breaks = age_group_names[whichLabels]) +
  facet_wrap(vars(cntry), ncol = 3) +
  labs(x = "Age group", y = "Mean happiness")
```



This is another quick visual, without any weighting.

```{r fig.height=15, fig.width=7}
theLambda <- 15
dat %>%
  select(agea, happy, cntry, pspwght) %>%
  na.omit %>%
  ggplot(aes(x = agea, y = happy, size = pspwght)) +
    geom_jitter(alpha = 0.2) + 
    scale_size(range = c(.8, 2), name="Weight") +
    geom_quantile(quantiles = c(.5),
                  method = "rqss",
                  formula = y ~ qss(x, lambda = theLambda),
                  size = 1.5,
                  alpha = 1,
                  col = "lightgreen") + 
    geom_quantile(quantiles = c(.1, .2, .8, .9),
                  method = "rqss",
                  formula = y ~ qss(x, lambda = theLambda),
                  size = .8,
                  alpha = .8,
                  col = "lightgreen") + 
    facet_wrap(vars(cntry), ncol = 3) +
    labs(x = "Age", y = "Happiness") +
    scale_x_continuous() + 
    scale_y_continuous() +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_rect(fill = "white"))

```


# Fit models

To get us started, let's fit a model to GB data.

```{r}
svy_gb <- dat_svy %>%
  filter(cntry == "GB")

mod_gb <- svy_gb %>%
  svyglm(happy ~ agea + I(agea^2),
         design = .,
         influence = T)
summary(mod_gb)
```

We can use tidy (in the broom package) to reshape this output to a handy tibble, which will be handy later for combining.

```{r}
tidy(mod_gb)
```



# Fit separate models to all countries

First step - package up the code above into a function taking a design object as input.

```{r}
happy_run <- function(the_design) {
  the_design %>%
  svyglm(happy ~ agea + I(agea^2),
         design = .) %>%
  tidy()
}
```

Test it out:

```{r}
dat_svy %>%
  filter(cntry == "GB") %>%
  happy_run()
```

Now we will apply to all the countries, so save their names.

```{r}
cs <- unique(dat_svy$variables$cntry)
table(cs)
```
Use `map_dfr` to fit the function to all countries data, producing a "long" tibble with all the coefficients.

```{r}
u_happy_coefs <- cs %>%
  map_dfr(function (x) 
            dat_svy %>%
            filter(cntry == x) %>%
            happy_run() %>%
            mutate(cntry = x, .before = "term")
  )
u_happy_coefs
```

Now plot the slopes for age and age-squared by country.

```{r}
threshold = 2

u_happy_coef_wide <- u_happy_coefs %>%
  select(cntry, term, statistic) %>%
  pivot_wider(names_from  = "term",
              values_from = "statistic") # the z-score

u_happy_coef_wide <- u_happy_coef_wide %>%
  mutate(class =
    case_when(
      agea < -threshold & `I(agea^2)` > threshold ~ "U-shaped",
      agea >  threshold & `I(agea^2)` < -threshold ~ "inverted-U",
      agea < -threshold &
        abs(`I(agea^2)`) < threshold              ~ "linear decrease",
      agea > threshold &
        abs(`I(agea^2)`) < threshold              ~ "linear increase",
      agea > 0 &
        abs(`I(agea^2)`) > threshold              ~ "quadratic increase",
      agea < 0 &
        abs(`I(agea^2)`) < -threshold              ~ "quadratic increase",
      abs(agea) < threshold & 
        abs(`I(agea^2)`) < threshold              ~ "constant",
      TRUE ~ "other"
    )
  )
    
u_happy_coef_wide %>%
  ggplot(aes(y = `I(agea^2)`, x = agea, label = cntry, colour = class)) +
    geom_text(check_overlap = TRUE) +
    labs(y = bquote(age^2~"(z statistic)"),
         x = "age (z statistic)") +
    geom_vline(xintercept = 0) +
    geom_hline(yintercept = 0) +  
    xlim(-4,4) +
    ylim(-4,4)
```




# To be continued...

The plan is:

1. Fit the U model to all data
2. Categorise them by whether the coefficients fit the U-pattern and calculate age of average worst happiness
3. Plot diagnostics for all countries
4. OMG maybe plot the slopes for x and x^2, coloured by country...



